Vagueness, Noise, and Communication

### 2014-09-30 paper/03_simulations.tex
results write-up sec com

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamics. The main question we should address is whether
diffusion and fitness-based replication reasonably interact, and if
so, whether the inclusion of diffusion has any noteworthy effects on
the evolving meaning of signals. Clearly, if $\impairment = 0$, the
\rdd reduces to the \rd. If $\impairment \rightarrow \infty$, the
diffusion component takes over and the \rdd reduces to the trivial
iterated diffusion that we looked at in the previous section. We will
show presently that for reasonable in-between levels of imprecision
$\impairment$, the \rdd leads to communicatively successful, yet vague
signal meaning. Moreover, suitable levels of imprecision can have
further accelerating and, therefore, unifying effects on meaning
evolution.

@@

Representative examples for resulting strategy pairs at stopping time
are given in
Figure~\ref{fig:example_strats}. Figure~\ref{fig:example_stratsA}
shows a strategy pair at stopping time with 90 states, tolerance
$\toler = 0.1$ and imprecision $\impairment = 0$. The latter means
that the trial was effectively a plain application of the
\rd. Noteworthily, the given sender strategy approximates a pure
sender strategy that crisply partitions the state space into
non-convex sets. The irregular shape of the receiver strategy shows
clearly that the pictured strategy pair has not yet reached a
dynamically stable state under the \rd. Indeed, the trial did not
converge in our technical sense, but was stopped after 200 rounds of
iterated \rdd. In contrast, the outcome of a trial with identical
parameters, except that imprecision was $\impairment = 0.1$, is shown
in Figure~\ref{fig:example_stratsB} had converged (in our technical
sense) after 98 rounds of iterated \rdd. The sender strategy shows a
smooth blending from one ``category'' to the other, and the receiver's
interpretations are rather extended curves, peaking at a central point
in the relevant ``categories''. What these examples show is that (i)
(speaker) strategies can approach non-convex strategies and that (ii)
inclusion of imprecision can lead to seemingly well-behaved, yet vague
strategies in the sense that we are after.

To explore further what happened in all of our 50 trials for each
parameter triple, we defined a small number of metrics that aim to
capture numerically how vague, communicatively efficient and generally
well-structured the resulting strategy pairs were. Entropy and
informativity capture (slightly different aspects of) the amount of
systematicity or regularity in a language. Voronoiness and convexity
are, respectively, gradient and categorical measures for the coherence
of linguistically formed categories.

\paragraph{Entropy.} This classic information-theoretic metric
captures the amount of uncertainty in a signaling strategy. Roughly
put, entropy of a strategy captures inverse distance from a pure
strategy in the present context. Vague languages will have higher
entropy than crisp ones.

The most natural way to define it is in terms of mixed strategies
rather than behavioral strategies.  Let us recall that mixed sender
(receiver) strategies are functions $\Smixed \in
\Delta(\Messgs^\States)$ ($\Rmixed \in (\Acts^\Messgs)$).  We can
define the entropy of a sender and receiver strategy as
\begin{align*}
  E(\Smixed) & = \sum_{\Spure \in \Messgs^\States} \Smixed(\Spure) \cdot
  \log(\Smixed(\Spure)) & 
  E(\Rmixed) & = \sum_{\Rpure \in \Acts^\Messgs} \Rmixed(\Rpure) \cdot \log(\Rmixed(\Rpure)) \,.
\end{align*} 
These measures are computationally expensive to calculate, since the
size of the domain over which the sum is computed grows exponentially
with the number of choice points.  Therefore, we
converted\footnote{See Appendix~\ref{sec:proofs}.}\todo{include
  this!?} the above definitions into equivalent metrics defined in
terms of behavioral strategies.  These are as follows:
\begin{align*}
  E(\Sstrat) = -\sum_{\state \in \States} \sum_{\messg \in \Messgs}
  \Sstrat(\messg \probbar \state) \cdot \log(\Sstrat(\messg \probbar
  \state)) 
\end{align*} 
\begin{align*}
  E(\Rstrat) = -\sum_{\messg \in \Messgs} \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot \log(\Rstrat(\act \probbar
  \messg)) \,. 
\end{align*}
Values obtained by these definitions are lower bounded by $0$ and
upper bounded by, respectively, $\log(|\Messgs^\States|) = |\States|
\cdot \log(|\Messgs|)$ and $\log(|\Acts^\Messgs|) = |\Messgs| \cdot
\log(|\Acts|)$. We work with normalized values.

The speaker strategies in Figures~\ref{fig:example_stratsA} and
\ref{fig:example_stratsB} have entropy $1.19e^{-5}$ and $0.21$,
respectively. The receiver strategies have respective entropies $0.43$
and $0.84$.

\paragraph{Informativity.} The quantity of information in a signal is
an old notion that also goes back to the start of information theory.
Skyrms~\citet[ch.~3]{Skyrms2010:Signals} discusses its use in the
context of signaling games. Intuitively speaking, informativity in
Skyrms' sense measures how revealing the message use is about which
state is actual. It therefore measures, in our given set-up with flat
priors, a sense of well-behavedness or regularity of a sender
strategy. Vague languages will have lower informativity than crisp
ones.

The main idea behind Skyrm's notion is that we can quantify the amount
of information about a state $\state$ in a message $\messg$ via the
relation between the probability of the state given the message
$\Pr(\state \,|\, \messg)$ and the unconditional probability of the
state $\Pr(\state)$.  Following Bayes' theorem, we can express
$\Pr(\state \,|\, \messg)$ as $\frac{\Pr(\state) \cdot \Pr(\messg
  \,|\, \state)}{\Pr(\messg)}$.  We then have $\Pr(\messg \,|\,
\state) = \Sstrat(\state,\messg)$ and $\Pr(\messg) =
\sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot
\Sstrat(\state^\prime,\messg)$.  Finally, we equate sender
infomativity with the average overall information about states in each
signal.  Following \citet[p.~36]{Skyrms2010:Signals}, and the
considerations above, the informativity about states conveyed by a
sender strategy is:
\begin{align*}
  I(\Sstrat) = \frac{1}{|\Messgs|} \sum_{\messg \in \Messgs}
  \sum_{\state \in \States} \frac{\Pr(\state) \cdot
    \Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States}
    \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \cdot \log
  \left(\frac{\Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States}
      \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \right)
  \,. 
\end{align*}
This measure is bounded between $0$ and $1$. The speaker strategies in
Figures~\ref{fig:example_stratsA} and \ref{fig:example_stratsB} have
informativity $0.69$ and $0.54$, respectively.

@@

Despite being vague, a language can have maximal Voronoiness according
to this gradient measure. For instance, the sender strategies in
Figures~\ref{fig:example_stratsA} and \ref{fig:example_stratsB} have
Voronoiness $0.91$ and $1$, respectively. The receiver strategies have
Voronoiness $0.78$ and $1$, respectively.

\paragraph{Convexity.} At least for speaker strategies, which converge
faster, it also makes sense to define a categorical measure of
convexity, that compensates for potential vagueness. To determine
whether a sender strategy $\Sstrat$ was convex despite possibly being
vague, we looked at the derived pure strategy $\Spure$ for which
$\Spure(\state) = \messg$ only if $\Sstrat(\state,\messg) =
\max_{\messg' \in \Messgs} \Sstrat(\state,\messg')$. If that $\Spure$
was convex, we also counted $\Sstrat$ as convex.

@@

\subsubsection{Results}
Figure~\ref{fig:MeanMetrics} shows the means of the recorded metrics,
together with their estimated 95\% confidence intervals, for each
triple of independent variables. Most trends are quite clear. Entropy
of sender and receiver strategies are increasing in imprecision
$\impairment$, as can be expected. Entropy of speaker strategies
hardly seems independent of tolerance levels and the number of states,
while there seems to be an effect of both parameters on the entropy of
the hearer strategy.

@@

Speaker informativity and expected utility are decreasing under
growing imprecision, again, as should be expected. There seem to be
mild effects of the number of states in the game, but hardly any in
terms of pragmatic tolerance.

Interestingly, voronoiness of both sender and receiver strategies are
appear to be non-decreasing under increasing impairment and peaking at
the maximal value. The only exception is the 6 states games and the
transition from $\impairment = 0$ to $\impairment = 0.05$. This
suggests that impairment can help evolve more regular, well-behaved
languages.

This impression is also supported by the proportion of converged and
convex trials, pictured in
Figure~\ref{fig:CategoricalMeasures}. Non-converged and non-convex
speaker strategies showed for several values of parameters in the
absence of impairment. Increasing impairment always assured
convergence and convexity.

@@

Summarizing so far, it seems that diffusion increases entropy and
vagueness, decreases informativity and expected utility, but it also
speeds up convergence to a convex and well-behaved outcome.

@@

This type of variability in evolutionary outcomes is also weeded out
by mild increases in impairment.

@@

    \includegraphics[width=\textwidth]{plots/AverageEUinGroups.pdf}

   \caption{Average normalized expected utility between evolved strategies within each triplet of independent parameter values.}
  \label{fig:AverageEUinGroups}
\end{figure}

>>>>

Now that we understand better what diffusion does, let us look at how
diffusion interacts with optimization of behavior, as described by the
replicator dynamics. The main question we should address is whether
diffusion and fitness-based replication reasonably interact, and if
so, whether the inclusion of diffusion has any noteworthy effects on
the evolving meaning of signals. Clearly, if $\impairment = 0$, the
\rdd reduces to the \rd. If $\impairment \rightarrow \infty$, the
diffusion component takes over and the \rdd reduces to the trivial
iterated diffusion that we looked at in the previous section. We will
show presently that for reasonable in-between levels of imprecision
$\impairment$, the \rdd leads to communicatively successful, yet vague
signal meaning. Suitable levels of imprecision can have further
accelerating and unifying effects on meaning evolution.

\subsubsection{Experimental set-up}

@@

Representative examples for resulting strategy pairs at stopping time
are given in
Figure~\ref{fig:example_strats}. Figure~\ref{fig:example_stratsA}
shows a strategy pair at stopping time with 90 states, tolerance
$\toler = 0.1$ and imprecision $\impairment = 0$. The latter means
that the trial was effectively a plain application of the
\rd. Noteworthily, the given sender strategy approximates a pure
sender strategy that crisply partitions the state space into
non-convex sets. The irregular shape of the receiver strategy shows
clearly that the pictured strategy pair has not yet reached a
dynamically stable state under the \rd. Indeed, the trial did not
converge in our technical sense, but was stopped after 200 rounds of
iterated \rdd. In contrast, the outcome of a trial with identical
parameters, except that imprecision was $\impairment = 0.1$, is shown
in Figure~\ref{fig:example_stratsB} had converged (in our technical
sense) after 99 rounds of iterated \rdd. The sender strategy shows a
smooth blending from one ``category'' to the other, and the receiver's
interpretations are rather extended curves, peaking at a central point
in the relevant ``categories''. What these examples show is that (i),
contrary to the limiting results of
\citet{Jager2007:The-Evolution-o,JagerMetzger2011:Voronoi-Languag},
(speaker) strategies can approach non-convex pure strategies and that
(ii) inclusion of imprecision can lead to seemingly well-behaved, yet
vague strategies in the sense that we are after.

\subsubsection{Dependent measures}

To explore further what happened in all of our 50 trials for each
parameter triple, we defined a small number of metrics that aim to
capture numerically how vague, communicatively efficient and generally
well-structured the resulting strategy pairs were. Entropy and
informativity capture (slightly different aspects of) the amount of
systematicity or regularity in a language. Voronoiness and convexity
are, respectively, gradient and categorical measures for the coherence
of linguistically formed categories. Expected utility measures the
communicative efficiency of evolved strategy pairs.

\paragraph{Entropy.} This classic information-theoretic metric
captures the amount of uncertainty in a signaling strategy. Roughly
put, entropy of a strategy captures inverse distance from a pure
strategy. The most natural way to define it is in terms of mixed
strategies rather than behavioral strategies.  Let us recall that
mixed sender (receiver) strategies are functions $\Smixed \in
\Delta(\Messgs^\States)$ ($\Rmixed \in (\Acts^\Messgs)$).  We can
define the entropy of a sender and receiver strategy as
\begin{align*}
  E(\Smixed) & = \sum_{\Spure \in \Messgs^\States} \Smixed(\Spure) \cdot
  \log(\Smixed(\Spure)) & 
  E(\Rmixed) & = \sum_{\Rpure \in \Acts^\Messgs} \Rmixed(\Rpure) \cdot \log(\Rmixed(\Rpure)) \,.
\end{align*} 
These measures are computationally expensive to calculate, since the
size of the domain over which the sum is computed grows exponentially
with the number of choice points.  Therefore, we
converted\footnote{See Appendix~\ref{sec:proofs}.}\todo{include
  this!?} the above definitions into equivalent metrics defined in
terms of behavioral strategies.  These are as follows:
\begin{align*}
  E(\Sstrat) = -\sum_{\state \in \States} \sum_{\messg \in \Messgs}
  \Sstrat(\messg \probbar \state) \cdot \log(\Sstrat(\messg \probbar
  \state)) 
\end{align*} 
\begin{align*}
  E(\Rstrat) = -\sum_{\messg \in \Messgs} \sum_{\act \in \Acts}
  \Rstrat(\act \probbar \messg) \cdot \log(\Rstrat(\act \probbar
  \messg)) \,. 
\end{align*}
Values obtained by these definitions are lower bounded by $0$ and
upper bounded by, respectively, $\log(|\Messgs^\States|) = |\States|
\cdot \log(|\Messgs|)$ and $\log(|\Acts^\Messgs|) = |\Messgs| \cdot
\log(|\Acts|)$. We work with normalized values.

The speaker strategies in Figures~\ref{fig:example_stratsA} and
\ref{fig:example_stratsB} have entropy $1.19e^{-5}$ and $0.21$,
respectively. The receiver strategies have respective entropies $0.43$
and $0.84$. In general, we expect that vague languages will have
higher entropy than crisp ones and that increasing imprecision will
also increase entropy, all else being equal.

\paragraph{Informativity.} The quantity of information in a signal is
an old notion that also goes back to the start of information theory.
\citet[ch.~3]{Skyrms2010:Signals} discusses its use in the context of
signaling games. Intuitively speaking, informativity in Skyrms' sense
measures how revealing the sender's signal use is about which state is
actual. It therefore measures, in our given set-up with flat priors, a
sense of well-behavedness or regularity of a sender strategy.

The main idea behind Skyrm's notion is that we can quantify the amount
of information about a state $\state$ in a message $\messg$ via the
relation between the probability of the state given the message
$\Pr(\state \,|\, \messg)$ and the unconditional probability of the
state $\Pr(\state)$.  Following Bayes' theorem, we can express
$\Pr(\state \,|\, \messg)$ as $\frac{\Pr(\state) \cdot \Pr(\messg
  \,|\, \state)}{\Pr(\messg)}$.  We then have $\Pr(\messg \,|\,
\state) = \Sstrat(\state,\messg)$ and $\Pr(\messg) =
\sum_{\state^\prime \in \States} \Pr(\state^\prime) \cdot
\Sstrat(\state^\prime,\messg)$.  Finally, we equate sender
infomativity with the average overall information about states in each
signal.  Following \citet[p.~36]{Skyrms2010:Signals}, and the
considerations above, the informativity about states conveyed by a
sender strategy is:
\begin{align*}
  I(\Sstrat) = \frac{1}{|\Messgs|} \sum_{\messg \in \Messgs}
  \sum_{\state \in \States} \frac{\Pr(\state) \cdot
    \Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States}
    \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \cdot \log
  \left(\frac{\Sstrat(\state,\messg)}{\sum_{\state^\prime \in \States}
      \Pr(\state^\prime) \cdot \Sstrat(\state^\prime,\messg)} \right)
  \,. 
\end{align*}
This measure is bounded between $0$ and $1$. The speaker strategies in
Figures~\ref{fig:example_stratsA} and \ref{fig:example_stratsB} have
informativity $0.69$ and $0.54$, respectively. In general, we expect
that vague languages will have lower informativity than crisp
ones and that increasing imprecision will decrease informativity.

@@

Despite being vague, a language can have maximal Voronoiness according
to this gradient measure. For instance, the sender strategies in
Figures~\ref{fig:example_stratsA} and \ref{fig:example_stratsB} have
Voronoiness $0.91$ and $1$, respectively. The receiver strategies have
Voronoiness $0.78$ and $1$, respectively. We do not \emph{prima facie}
expect a conceptual connection between vagueness and Voronoiness, and
so do not predict a clear effect of imprecision on Voronoiness either.

\paragraph{Convexity.} At least for speaker strategies, which converge
faster, it also makes sense to define a categorical measure of
convexity, that compensates for potential vagueness. To determine
whether a sender strategy $\Sstrat$ was convex despite possibly being
vague, we looked at the derived pure strategy $\Spure$ for which
$\Spure(\state) = \messg$ only if $\messg =
\argmax_{\messg' \in \Messgs} \Sstrat(\state,\messg')$. If that $\Spure$
was convex, we also counted $\Sstrat$ as convex. The sender strategy
in Figure~\ref{fig:example_stratsA} is not convex, while the one in
Figure~\ref{fig:example_stratsB} is. Again, we do not expect a
conceptual relation between vagueness and imprecision, on the one
hand, and convexity on the other.

@@

\subsubsection{Results}
Figure~\ref{fig:MeanMetrics} shows the means of the recorded metrics,
together with their estimated 95\% confidence intervals, for each
triple of independent variables. Most trends are quite clear, and not
unexpected. Entropy of sender and receiver strategies are increasing
in imprecision. Entropy of speaker strategies seems independent of
tolerance levels and the number of states, while there seems to be an
effect of both parameters on the entropy of the hearer
strategy. Speaker informativity and expected utility are decreasing
under growing imprecision. There seem to be mild effects of the number
of states in the game, but hardly any in terms of pragmatic
tolerance. Interestingly, Voronoiness of both sender and receiver
strategies appear to be non-decreasing under increasing impairment and
peaking at the maximal value. The only exception is the 6 states games
and the transition from $\impairment = 0$ to $\impairment =
0.05$. Expected utility of evolved languages mostly decreases in
increasing imprecision, with the sole exception of a rise for $\toler
= 0.2$ and $ns = 6$ at $\impairment = 0.1$. Still, normalized expected
utility of evolved strategy pairs under mild values of imprecision are
not necessarily much worse than their respective perfect precision
\rd-analogues.

@@

Taken together, these findings suggest the following effect of
imprecise perception and diffusion of behavior. Mild forms of
perceptual imprecision lead to slightly less communicative efficiency
in the evolving strategy pairs, more vagueness and more regular,
well-behaved languages. This impression is also supported by the
proportion of converged and convex trials, pictured in
Figure~\ref{fig:CategoricalMeasures}. Non-converged and non-convex
speaker strategies showed for several values of parameters in the
absence of impairment. Increasing impairment always assured
convergence and convexity: perceptual imprecision speeds up
convergence to a convex and regular language.

@@

@@

Interestingly, this type of variability in evolutionary outcomes is
also weeded out by mild increases in impairment. To see this, we
calculated the average distance between evolved sender strategies
within each group of trials that had identical independent
parameters. We determined the distance between sender strategies
$\Sstrat$ and $\Sstrat'$ as the average Hellinger distance between
probability distributions $\Sstrat(\state)$ and $\Sstrat'(\state)$
over messages at each choice point $\state$:
\begin{align*}
  \text{HD}(\Sstrat,\Sstrat') & = \frac{1}{{\card{\States} \cdot
     \sqrt{2}}} \cdot  \sum_{\state \in \States} 
 \sqrt{\sum_{\messg \in  \Messgs}
         \left ( \sqrt{\Sstrat(\state,\messg)} -
         \sqrt{\Sstrat'(\state,\messg)} \right )^2 }\,.
      \\
\end{align*}
To compensate for the arbitrariness of message use, we set the
distance between strategies $\Sstrat$ and $\Sstrat'$ to be the maximum
of $\text{HD}(\Sstrat,\Sstrat')$ and $\text{HD}(\Sstrat^*,\Sstrat')$
where $\Sstrat^*$ is $Sstrat$ with reversed message indices. The
average distances between sender strategies in this sense is plotted
in Figure~\ref{fig:AverageEUinGroups}. This clearly shows that with
increasing imprecision, the sender strategies evolving under \rdd were
exactly alike, modulo which message was used for which part of the
unit interval. In other words, perceptual imprecision clearly unifies
evolutionary outcomes, and in fact guarantees sender strategies that
are not only convex, but also maximally efficient in that they induce
a vague category split exactly in the middle of the unit interval.

@@

    \includegraphics[width=\textwidth]{plots/WithinGroupMeasures.pdf}

    \caption{Mean group measures as defined in the main text. For each
      group, i.e., triplet of independent variables, the plot shows
      (i) average normalized expected utility when playing against an
      arbitrary other language from the group, (ii) the mean distance
      between evolved strategies in the group.}
  \label{fig:AverageEUinGroups}
\end{figure}

This unifying property of perceptual imprecision could be considered
an evolutionary beneficial side-effect. When we look at the normalized
average expected utility that each evolved language (i.e., pair of
sender and receiver strategy at stopping time) scored when playing
against an arbitrary other language of the same group in
Figure~\ref{fig:MoreExample_strats}, we see that higher imprecision
can lead to higher such average expected utility. What this means is
that, imprecision might decrease the communicative efficiency of
individual languages, it increases the conceptual coherence and
communicative success between independently evolving strategies. It is
almost as if mere confusability of states imposes a regularity
constraint on evolving linguistic meanings.
